Approximate bayesian computation in large scale structure modeling : constraining the galaxy-halo connection

### 2016-01-25 halo/text/abc.tex

\documentclass[12pt, preprint]{aastex}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

@@

\subsection{Forward model of the large scale galaxy populations}\label{sec:halo}
\subsubsection{halo model}
Dark matter halos are biased tracers of the underlying matter distribution of the universe. The halo model postulates that galaxies reside in dark matter halos. Halo Occupation distribution---HOD---is a powerful tool for description of galaxy--halo connection. HOD provides a probabilistic prescription for how galaxies populate halos. According to HOD modeling, the number of galaxies that populate a dark matter halo is dictated by the host halo mass $M_h$. The probability of a dark matter halo of mass $M_h$ being populated by $N_g$ galaxies---$P(N_g|M_h)$--- is a Poisson distribution, and is separable to two parts 
\beq
P(N_g|M_h) = P(N_\mathrm{cen}|M_h) + P(N_\mathrm{sat}|M_h),
\eeq
where $N_\mathrm{cen}$ is the number of central galaxies, and $N_\mathrm{sat}$ is the number of satellites.

@@

talk about are results ...

talk about future directions ...
\bibliographystyle{yahapj}
\bibliography{ccppabc}
\end{document}

>>>>

ss[12pt, preprint]{aastex}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

@@

\subsection{Forward model of the large scale galaxy populations}\label{sec:halo}
\subsubsection{halo occupation modeling}

\newcommand{\lcdm}{\Lambda {\rm CDM}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\mean}[2]{\left\langle#1 \vert {#2}\right\rangle}

%%% HALO OCCUPATION DISTRIBUTION %%%
\newcommand{\ngal}{N_{\mathrm{g}}}
\newcommand{\nsat}{N_\mathrm{s}}
\newcommand{\ncen}{N_\mathrm{c}}
\newcommand{\pnm}[2]{P(#1|#2)}

\newcommand{\mhalo}{M_{\rm h}}
\newcommand{\mvir}{M_\mathrm{vir}}

\newcommand{\dndmvir}{\frac{\dd n}{\dd\mvir}}
\newcommand{\dndmhalo}{\frac{\dd n}{\dd\mhalo}}
\newcommand{\dndmvirprime}{\frac{\dd n}{\dd\mvir'}}

%%% CLUSTERING STATISTICS %%%
\newcommand{\xigg}{\xi_{\mathrm{gg}}}
\newcommand{\xihh}{\xi_{\mathrm{hh}}}
\newcommand{\xiggr}{\xi_{\mathrm{gg}}(r)}
\newcommand{\xiggroneh}{\xi^{1h}_{\mathrm{gg}}(r)}
\newcommand{\xiggrtwoh}{\xi^{2h}_{\mathrm{gg}}(r)}
\newcommand{\ngalaxy}{\bar{n}_{\mathrm{g}}}

One of the most well-studied statistics of the galaxy distribution 
is $\xigg,$ the galaxy-galaxy correlation function. Mathematically, $\xigg$ encodes the
excess probability that a pair of galaxies in a sample are 
separated by three-dimensional distance $r.$ 
Galaxy clustering is straightforward to measure observationally, 
and it is a highly sensitive probe of both the physics of galaxy evolution 
as well as the fundamental parameters of $\lcdm$ cosmology.

The assumption that galaxies reside in dark matter halos is the bedrock underlying 
all contemporary theoretical predictions for galaxy clustering. Since the 
clustering of dark matter halos can be predicted to high precision with N-body simulations, 
then knowledge of how galaxies populate and are distributed within halos 
is sufficient to predict $\xigg.$ Under this assumption, $\xigg$ can be formally decomposed
into a ``one-halo term" and a ``two-halo term'', 
referring to galaxy pairs that reside in the same halo vs. distinct halos, respectively. 
The two-halo term of $\xigg$ is a pair-number weighted average of $\xihh,$ 
the halo-halo correlation function, whereas the one-halo term is the pair-number 
weighted average of the intra-halo distribution of galaxies. 
Models for the galaxy-halo connection aiming to predict $\xigg$ must therefore 
provide theoretical means by which to calculate these weights and spatial distributions.

The Halo Occupation Distribution (HOD) is one of the most widely used 
 approaches to characterizing the galaxy-halo connection. 
The central quantity In the HOD is $\pnm{\ngal}{\mhalo},$ the probability that a halo of mass
$\mhalo$ hosts $\ngal$ galaxies. The first two moments of $\pnm$ contain the 
necessary information to calculate the one- and two-halo terms of the galaxy correlation function:

\begin{eqnarray}
\label{eq:onehaloterm}
1+\xiggroneh \simeq \frac{1}{4\pi{}r^{2}\ngalaxy^{2}}\int\dd\mhalo\dndmhalo\Xi_{\rm gg}(r|\mhalo) \times \mean{\ngal(\ngal-1)}{\mhalo},
\end{eqnarray} and

\begin{eqnarray}
\label{eq:twohaloterm}
\xiggrtwoh \simeq \xi_{\mathrm{mm}}(r)\left[\frac{1}{\ngalaxy}\int\dd\mhalo\dndmhalo \mean{\ngal}{\mhalo}b_{\mathrm{h}}(\mhalo)\right]^{2}
\end{eqnarray}

In Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}), $\ngalaxy$
is the galaxy number density,
$\dd\mathrm{n}/\dd\mhalo$ is the halo mass function, 
the spatial bias of dark matter halos is $b_{\mathrm{h}}(\mhalo),$ and
$\xi_{\rm mm}$ is the correlation function of dark matter.  
If we represent the spherically symmetric intra-halo distribution of galaxies 
by a unit-normalized $n_{\rm g}(r),$
then the quantity $\Xi_{\rm gg}(r)$ appearing in the above two equations 
is the convolution of $n_{\rm g}(r)$ with itself.

Employing Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}) to 
calculate $\xigg(r)$ is the most common technical method used in HOD 
studies of galaxy clustering. 
However, this technique requires appeal to many simplifying assumptions 
in addition to the supposition that galaxies reside in halos. For example, 
Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}) presume that 
the intra-halo galaxy distribution is spherically symmetric, and 
that the halo bias $b_{\mathrm{h}}(\mhalo)$ has no radial dependence. 
Moreover, these equations does not take halo exclusion into account. 
We refer the interested reader to \citet{cooray02,mo_vdb_white10,vdBosch13}, 
and references therein, for further information.

In addition to the above, in the vast majority of HOD studies the actual calculation of 
Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}) relies on further assumption still. 
The quantities $b_{\mathrm{h}}(\mhalo),$ $\xi_{\mathrm{mm}}(r)$ and $\dndmhalo$ 
are commonly replaced by analytical fitting functions that have been calibrated 
against a suite of N-body simulations \citep[e.g.,][]{sheth_tormen01,smith03,tinker05,tinker10}.   
And yet, it is well established that the precision of these fitting functions is 
limited to the $\sim10\%$ level \citep[e.g.][]{tinker08}.

\todo{APH: Ok, so the above level of detail may be overkill for this paper. I figured that part of the argument we'd make is this: it is now time to start taking the precision of our predictions seriously, the ten-percent game is over. ABC is a big contribution to going beyond ten percent, and now that Halotools makes it possible to circumvent the need for these fitting functions, this opens up new limiting factors such as the hard Gaussianity assumption. So I see the above argument as part of the argument for using ABC. However, I'm not married to any of the above text, so just let me know if you want me to scrap some of it. Otherwise, I will next pick up from here and start talking about how Halotools relaxes the reliance on these approximations, and I will make explicit the point about how this new level of precision motivates the exploration of ABC.}

\todo{MJ: I think all of the text written above is necessary for making a case for simulation-based inference in large-scale structure modeling. So I'm in favor of keeping it for now.}

Dark matter halos are biased tracers of the underlying matter distribution of the universe. The halo model postulates that galaxies reside in dark matter halos. Halo Occupation distribution---HOD---is a powerful tool for description of galaxy--halo connection. HOD provides a probabilistic prescription for how galaxies populate halos. According to HOD modeling, the number of galaxies that populate a dark matter halo is dictated by the host halo mass $M_h$. The probability of a dark matter halo of mass $M_h$ being populated by $N_g$ galaxies---$P(N_g|M_h)$--- is a Poisson distribution, and is separable to two parts 
\beq
P(N_g|M_h) = P(N_\mathrm{cen}|M_h) + P(N_\mathrm{sat}|M_h),
\eeq
where $N_\mathrm{cen}$ is the number of central galaxies, and $N_\mathrm{sat}$ is the number of satellites.

@@

talk about are results ...
talk about future directions ...

%%%%%%%%%% FIGURES %%%%%%%%%%

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{xi_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy two-point correlation function signal in Fig. 1, where the covariance matrix is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{gmf_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy group multiplicity function signal in Fig. 2, where the covariance matrix
is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_sigma.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior predictions
for the two-point correlation function of galaxies. The black lines show the data-points.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_res_sigma3.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior model prediction residuals
for the two-point correlation function of galaxies.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{gmf3.png}
\caption{{\bf Top Panel:}1-$\sigma$(dark-shaded) and 2-$\sigma$ ABC posterior model prediction 
for the group multiplicity function of galaxies in the mock catalog.{\bf Bottom Panel:}Same as the top panel, but showing the residuals.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\section*{acknowledgments}

We would like to thank Mike Blanton, Boris Leidstadt, Micheal Schneider, Uros Seljak, Ross Fadely, Jeremy Tinker, Alex Malz, Patrick McDonald, Dan Foreman-Mackey, for productive and insightful discussions. 
MJV was supported by NSF grant AST-1517237. CHH was supported by James Arthur Graduate Award Grant.
DWH was supported by NSF (grants IIS-1124794 and AST-1517237), NASA (grant NNX12AI50G), and the Moore-Sloan Data Science Environment at NYU.

\bibliographystyle{yahapj}
\bibliography{ccppabc}
\end{document}
\documentclass[12pt, preprint]{aastex}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

\subsection{Forward model of the large scale galaxy populations}\label{sec:halo}
\subsubsection{halo occupation modeling}

\newcommand{\lcdm}{\Lambda {\rm CDM}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\mean}[2]{\left\langle#1 \vert {#2}\right\rangle}

%%% HALO OCCUPATION DISTRIBUTION %%%
\newcommand{\ngal}{N_{\mathrm{g}}}
\newcommand{\nsat}{N_\mathrm{s}}
\newcommand{\ncen}{N_\mathrm{c}}
\newcommand{\pnm}[2]{P(#1|#2)}

\newcommand{\mhalo}{M_{\rm h}}
\newcommand{\mvir}{M_\mathrm{vir}}

\newcommand{\dndmvir}{\frac{\dd n}{\dd\mvir}}
\newcommand{\dndmhalo}{\frac{\dd n}{\dd\mhalo}}
\newcommand{\dndmvirprime}{\frac{\dd n}{\dd\mvir'}}

%%% CLUSTERING STATISTICS %%%
\newcommand{\xigg}{\xi_{\mathrm{gg}}}
\newcommand{\xihh}{\xi_{\mathrm{hh}}}
\newcommand{\xiggr}{\xi_{\mathrm{gg}}(r)}
\newcommand{\xiggroneh}{\xi^{1h}_{\mathrm{gg}}(r)}
\newcommand{\xiggrtwoh}{\xi^{2h}_{\mathrm{gg}}(r)}
\newcommand{\ngalaxy}{\bar{n}_{\mathrm{g}}}

One of the most well-studied statistics of the galaxy distribution 
is $\xigg,$ the galaxy-galaxy correlation function. Mathematically, $\xigg$ encodes the
excess probability that a pair of galaxies in a sample are 
separated by three-dimensional distance $r.$ 
Galaxy clustering is straightforward to measure observationally, 
and it is a highly sensitive probe of both the physics of galaxy evolution 
as well as the fundamental parameters of $\lcdm$ cosmology.

The assumption that galaxies reside in dark matter halos is the bedrock underlying 
all contemporary theoretical predictions for galaxy clustering. Since the 
clustering of dark matter halos can be predicted to high precision with N-body simulations, 
then knowledge of how galaxies populate and are distributed within halos 
is sufficient to predict $\xigg.$ Under this assumption, $\xigg$ can be formally decomposed
into a ``one-halo term" and a ``two-halo term'', 
referring to galaxy pairs that reside in the same halo vs. distinct halos, respectively. 
The two-halo term of $\xigg$ is a pair-number weighted average of $\xihh,$ 
the halo-halo correlation function, whereas the one-halo term is the pair-number 
weighted average of the intra-halo distribution of galaxies. 
Models for the galaxy-halo connection aiming to predict $\xigg$ must therefore 
provide theoretical means by which to calculate these weights and spatial distributions.

The Halo Occupation Distribution (HOD) is one of the most widely used 
 approaches to characterizing the galaxy-halo connection. 
The central quantity In the HOD is $\pnm{\ngal}{\mhalo},$ the probability that a halo of mass
$\mhalo$ hosts $\ngal$ galaxies. The first two moments of $\pnm$ contain the 
necessary information to calculate the one- and two-halo terms of the galaxy correlation function:

\begin{eqnarray}
\label{eq:onehaloterm}
1+\xiggroneh \simeq \frac{1}{4\pi{}r^{2}\ngalaxy^{2}}\int\dd\mhalo\dndmhalo\Xi_{\rm gg}(r|\mhalo) \times \mean{\ngal(\ngal-1)}{\mhalo},
\end{eqnarray} and

\begin{eqnarray}
\label{eq:twohaloterm}
\xiggrtwoh \simeq \xi_{\mathrm{mm}}(r)\left[\frac{1}{\ngalaxy}\int\dd\mhalo\dndmhalo \mean{\ngal}{\mhalo}b_{\mathrm{h}}(\mhalo)\right]^{2}
\end{eqnarray}

In Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}), $\ngalaxy$
is the galaxy number density,
$\dd\mathrm{n}/\dd\mhalo$ is the halo mass function, 
the spatial bias of dark matter halos is $b_{\mathrm{h}}(\mhalo),$ and
$\xi_{\rm mm}$ is the correlation function of dark matter.  
If we represent the spherically symmetric intra-halo distribution of galaxies 
by a unit-normalized $n_{\rm g}(r),$
then the quantity $\Xi_{\rm gg}(r)$ appearing in the above two equations 
is the convolution of $n_{\rm g}(r)$ with itself.

Employing Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}) to 
calculate $\xigg(r)$ is the most common technical method used in HOD 
studies of galaxy clustering. 
However, this technique requires appeal to many simplifying assumptions 
in addition to the supposition that galaxies reside in halos. For example, 
Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}) presume that 
the intra-halo galaxy distribution is spherically symmetric, and 
that the halo bias $b_{\mathrm{h}}(\mhalo)$ has no radial dependence. 
Moreover, these equations does not take halo exclusion into account. 
We refer the interested reader to \citet{cooray02,mo_vdb_white10,vdBosch13}, 
and references therein, for further information.

In addition to the above, in the vast majority of HOD studies the actual calculation of 
Eqs.~(\ref{eq:onehaloterm}) and (\ref{eq:twohaloterm}) relies on further assumption still. 
The quantities $b_{\mathrm{h}}(\mhalo),$ $\xi_{\mathrm{mm}}(r)$ and $\dndmhalo$ 
are commonly replaced by analytical fitting functions that have been calibrated 
against a suite of N-body simulations \citep[e.g.,][]{sheth_tormen01,smith03,tinker05,tinker10}.   
And yet, it is well established that the precision of these fitting functions is 
limited to the $\sim10\%$ level \citep[e.g.][]{tinker08}.

\todo{APH: Ok, so the above level of detail may be overkill for this paper. I figured that part of the argument we'd make is this: it is now time to start taking the precision of our predictions seriously, the ten-percent game is over. ABC is a big contribution to going beyond ten percent, and now that Halotools makes it possible to circumvent the need for these fitting functions, this opens up new limiting factors such as the hard Gaussianity assumption. So I see the above argument as part of the argument for using ABC. However, I'm not married to any of the above text, so just let me know if you want me to scrap some of it. Otherwise, I will next pick up from here and start talking about how Halotools relaxes the reliance on these approximations, and I will make explicit the point about how this new level of precision motivates the exploration of ABC.}

\todo{MJ: I think all of the text written above is necessary for making a case for simulation-based inference in large-scale structure modeling. So I'm in favor of keeping it for now.}

Dark matter halos are biased tracers of the underlying matter distribution of the universe. The halo model postulates that galaxies reside in dark matter halos. Halo Occupation distribution---HOD---is a powerful tool for description of galaxy--halo connection. HOD provides a probabilistic prescription for how galaxies populate halos. According to HOD modeling, the number of galaxies that populate a dark matter halo is dictated by the host halo mass $M_h$. The probability of a dark matter halo of mass $M_h$ being populated by $N_g$ galaxies---$P(N_g|M_h)$--- is a Poisson distribution, and is separable to two parts 
\beq
P(N_g|M_h) = P(N_\mathrm{cen}|M_h) + P(N_\mathrm{sat}|M_h),
\eeq
where $N_\mathrm{cen}$ is the number of central galaxies, and $N_\mathrm{sat}$ is the number of satellites.

talk about are results ...
talk about future directions ...

%%%%%%%%%% FIGURES %%%%%%%%%%

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{xi_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy two-point correlation function signal in Fig. 1, where the covariance matrix is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{gmf_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy group multiplicity function signal in Fig. 2, where the covariance matrix
is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_sigma.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior predictions
for the two-point correlation function of galaxies. The black lines show the data-points.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_res_sigma3.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior model prediction residuals
for the two-point correlation function of galaxies.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{gmf3.png}
\caption{{\bf Top Panel:}1-$\sigma$(dark-shaded) and 2-$\sigma$ ABC posterior model prediction 
for the group multiplicity function of galaxies in the mock catalog.{\bf Bottom Panel:}Same as the top panel, but showing the residuals.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\section*{acknowledgments}

We would like to thank Mike Blanton, Boris Leidstadt, Micheal Schneider, Uros Seljak, Ross Fadely, Jeremy Tinker, Alex Malz, Patrick McDonald, Dan Foreman-Mackey, for productive and insightful discussions. 
MJV was supported by NSF grant AST-1517237. CHH was supported by James Arthur Graduate Award Grant.
DWH was supported by NSF (grants IIS-1124794 and AST-1517237), NASA (grant NNX12AI50G), and the Moore-Sloan Data Science Environment at NYU.

\bibliographystyle{yahapj}
\bibliography{ccppabc}
\end{document}
